---
title: "Ch04"
author: "VincentAlegrete"
date: "2024-12-24"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
register_knitr_engine(override = TRUE)
```

We will do something equivalent to the t-test using Stan. Using the data generated from the following R code, we want to decide whether the mean parameter $\mu_1$ in group one (<tt>Y1</tt>) differs from $\mu_2$ in group two (<tt>Y2</tt>) by computing $\text{Prob}[\mu_1 < \mu_2]$, which is the probability of $\mu_1 < \mu_2$.

```{r}
set.seed(123)
N1 <- 30
N2 <- 20
Y1 <- rnorm(n=N1, mean=0, sd=5)
Y2 <- rnorm(n=N2, mean=1, sd=4)
```

(1) Visualize the data from these two groups so that we can intuitively see whether the difference exists between them.

<i>Solution</i> Boxplots for $Y_1$ and $Y_2$ suggest that the distribution centers and spreads are different.

```{r}
boxplot(Y1, Y2, names=c("Y1", "Y2"), horizontal=TRUE)
```

(2) Write a model formula with the assumption that these two groups have the same SD. This corresponds to the Students’ t-test.

<i>Solution</i>

With the assumption that $\text{Var}(Y_1)=\text{Var}(Y_2)$, then our likelihoods (under flat priors for $\mu_1,\mu_2,\sigma$) are
$$Y_1|\mu_1,\sigma\sim N(\mu_1,\sigma^2)$$
$$Y_2|\mu_2,\sigma\sim N(\mu_2,\sigma^2)$$
We will build posterior distributions for both means in the next two subproblems.

(3) Create the model file of (2) in Stan and estimate the parameters. Don’t use <tt>generated quantities</tt> block here yet, because in the next (4) we will be practicing how to make use of draws from R or Python.

```{stan output.var="model3"}
data {
  int N1;
  int N2;
  vector[N1] Y1;
  vector[N2] Y2;
}

parameters {
  real mu1;
  real mu2;
  real<lower=0> sigma;
}

model {
  Y1 ~ normal(mu1, sigma);
  Y2 ~ normal(mu2, sigma);
}
```


(4) Compute Prob[μ1 < μ2] from the obtained draws using R or Python (hint: we can count how many times the event μ1 < μ2 occurs in the entire draws, and divide this quantity by the total number of draws).

<i>Solution</i> We run the model in STAN, generate our posterior draws, and display the model summary.

```{r}
library(cmdstanr)
data <- list(N1=N1, N2=N2, Y1=Y1, Y2=Y2)
# model3 <- cmdstan_model(stan_file='model3.stan')
fit <- model3$sample(data=data, seed=200)
fit$cmdstan_summary()
```

Now, we store the posterior draws into a data frame and compute $\text{Prob}[\mu_1 < \mu_2]$ as relative frequency:

```{r}
draws <- data.frame(fit$draws())
prob = (sum(draws$X1.mu1 < draws$X1.mu2) +
        sum(draws$X2.mu1 < draws$X2.mu2) +
        sum(draws$X3.mu1 < draws$X3.mu2) +
        sum(draws$X4.mu1 < draws$X4.mu2) ) / 4000
prob
```

So the posterior probability that $\mu_1 < \mu_2$ is approximately $92.35\%$. 

(5) Write a model formula with the assumption that the two SDs are different. This is equivalent to the Welch’s t-test. Similarly, compute Prob[μ1 < μ2]. 